---
title: "UFO EDA"
format: html
editor: visual
---

Simple exploration of the UFO data stored on Databricks

```{r setup}
# Packages ----
library(sparklyr)
library(pysparklyr)
library(tidyverse)
library(dbplyr)
library(DBI)

# Plot themes ----
theme_set(theme_bw())
```

We can connect to Databricks using either `sparklyr` or `odbc`:

```{r sparklyr-connection}
# Replace cluster_id with the appropriate value for your Databricks workspace
sc <- spark_connect(
    cluster_id = Sys.getenv("DATABRICKS_CLUSTER_ID"),
    method = "databricks_connect"
)

```

```{r odbc-connection}
# Without the new odbc::databricks() function
con <- dbConnect(
  odbc::odbc(),
  driver = "databricks",
  Host = Sys.getenv("DATABRICKS_HOST"),
  Port = 443,
  HTTPPath = Sys.getenv("DATABRICKS_HTTP_PATH"),
  Catalog = "samples",
  SSL = 1,
  ThriftTransport = 2,
  AuthMech = 3,
  UID = "token",
  PWD = Sys.getenv("DATABRICKS_TOKEN")
)

dbDisconnect(con)

# With the new odbc::databricks() function
con <- dbConnect(
  odbc::databricks(),
  HTTPPath  = Sys.getenv("DATABRICKS_HTTP_PATH"),
  timeout   = 10
)
```

Once connected, we can create a reference to the `nuforc` table. Functions executed against this reference will result in queries executed on Databricks.

```{r data}
nuforc <- tbl(sc, in_catalog("demos", "nuforc", "nuforc_reports"))
head(nuforc) |> 
  show_query()
```

## EDA

-   Frequency of sightings over time (by day)

    ```{r}
    ufo_by_day <- nuforc |> 
      filter(!is.na(date_time)) |> 
      count(date = as.character(as.Date(date_time))) |> 
      collect() |> 
      mutate(date = as.Date(date),
             month = month(date),
             year = year(date))

    ufo_by_day |> 
      ggplot(aes(x = month, y = n)) +
      facet_wrap(~year) +
      geom_col() +
      theme(axis.text.x = element_blank(),
            axis.text.y = element_blank()) +
      labs(title = "UFO Sightings by Month",
           x = "",
           y = "")
    ```

-   Aggregate UFO sightings over time

    ```{r}
    cummulative_ufo <- nuforc |> 
      filter(!is.na(date_time)) |> 
      count(date = as.Date(date_time)) |> 
      arrange(date) |> 
      mutate(cummulative_sightings = cumsum(n),
             date = as.character(date)) |> 
      collect()

    cummulative_ufo |> 
      ggplot(aes(x = as.Date(date), y = cummulative_sightings)) +
      geom_line() +
      scale_y_continuous(labels = scales::label_comma()) + 
      labs(x = "Date", y = "UFO Sightings", title = "Cummulative UFO Sightings")
    ```

-   How many sightings by Country?

    ```{r}
    (country_count <- nuforc |> 
        filter(!is.na(country)) |> 
        count(country, sort = TRUE) |> 
        collect())
    ```

-   How many sightings by State (USA)?

    ```{r}
    (state_count <- nuforc |> 
        filter(country == "USA", !is.na(state)) |> 
        count(state, sort = TRUE) |> 
        collect())
    ```

    ```{r}
    state_count |> 
      filter(n > 10) |> 
      ggplot(aes(x = fct_reorder(state, n), y = n)) +
      geom_col() +
      scale_y_continuous(labels = scales::label_comma()) +
      labs(x = "State", y = "UFO Sightings", title = "UFO Sightings by State (USA)") +
      coord_flip()
    ```

## Close Connections

```{r}
dbDisconnect(con)
spark_disconnect(sc)
```
